---
title: "lab_02"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most?

**A1: 32 departments have offered courses with "Climate" in the title with the Atmospheric and Oceanic Science department having the most (93 courses)**

```{r}

climate_courses <- umd_courses |> 
  group_by(department) |> 
  filter(str_detect(title, "Climate") & seats >= 1) |> 
  summarize(word_climate = n()) |> 
  arrange(desc(word_climate))
```

**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why?

**A2: Anthropology is the most interesting to me because I wasn't expecting it to be the second most and above other STEM departments such as Chemistry or Agricultural and Resource Economics**

**Q3** Which departments offered the most courses in the Fall 2024 term? And which departments in that term have at least one syllabus posted? Answer both of those questions and then compare the top five departments - how are they different?

**A3: Business & Management, Economics, School of Music, Information Studies, and Engineering, Professional Matters offered the most courses in Fall 2024; 134 courses in that term had at least one syllabus posted; and the top five departments are different because Business & Management have double the amount of courses in this term compared to the other four that average 50 courses**

```{r}

fall_24 <- umd_courses |> 
  group_by(department) |>
  filter(term == 202408 & syllabus_count >= 1) |> 
  summarise(course_count = n()) |> 
    arrange(desc(course_count))
```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as a new dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name, and we're not looking in payee_name). Write a sentence describing the pattern of spending over time.

**A4: The summer months leading up to Election Day saw the most donations from Wes Moore's committee, which could indicate them trying to boost his campaign up in the polls**

```{r}
  
  wes_moore <- maryland_expenses |>
    mutate(expenditure_date = mdy(expenditure_date)) |>
    mutate(month = floor_date(expenditure_date, "month")) |>
    filter(payee_type == "Candidate Committee" & str_detect(committee_name, "Wes For")) |>
    select(payee_type, committee_name, month, amount) |>
    group_by(month) |>
    summarize(total_donations = sum(amount)) |>
    arrange(month)
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for all others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).

**A5: The total amount of money spent in-state is \$51,801, out-of-state is \$29,281 and the amount with an unrecorded state is \$16,830**

```{r}

in_out_spending <- maryland_expenses |> 
  mutate(
    state = case_when(
      str_detect(address, " Maryland ") ~ "MD",
      str_detect(address, " California ") ~ "CA",
      str_detect(address, " Washington ") ~ "WA",
      str_detect(address, " Louisiana ") ~ "LA",
      str_detect(address, " Florida ") ~ "FL",
      str_detect(address, " North Carolina ") ~ "NC",
      str_detect(address, " Massachusetts ") ~ "MA",
      str_detect(address, " West Virginia ") ~ "WV",
      str_detect(address, " Virginia ") ~ "VA",
      .default = NA)) |> 
  mutate(spending_location = case_when(
    str_detect(state, "MD") ~ "in",
    str_detect(state, "CA") ~ "out",
    str_detect(state, "WA") ~ "out", 
    str_detect(state, "LA") ~ "out",
    str_detect(state, "FL") ~ "out",
    str_detect(state, "MA") ~ "out", 
    str_detect(state, "WV") ~ "out", 
    str_detect(state, "VA") ~ "out",
    str_detect(state, "NA") ~ "NA")) |>
  group_by(spending_location) |>
  summarize(total_amount = n())
```

**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab.

**A6: Who did Google donate to the most? Answer: Blegay Wala**

```{r}

google_donations <- maryland_expenses |> 
  filter(payee_name == "Google") |> 
  select(payee_name, amount, committee_name) |>
  group_by(committee_name) |>
  summarize(total_donations = sum(amount)) |>
  arrange(desc(total_donations))
```
