---
title: "lab_03"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

library(janitor)
library(tidyverse)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
payments <- read.csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv") 
md_grants <- read.csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")

cleaned_md_grants <- md_grants |>
  clean_names() 

cleaned_payments <- payments |> 
  clean_names()
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

**A1: We found 64 records**

```{r}

pg_schools <- cleaned_payments |>
  filter(str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOL") | 
           str_detect(vendor_name, "PRINCE GEORGE'S COUNTY PUBLIC SCHOOL") |
           str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL") | 
           str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGE") | 
           str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGE'S COUNTY") | 
           str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD") |
           str_detect(vendor_name, "LITERACY COUNCIL OF PG COUNTY MARYLAND" )) 
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

**A2:** **2024 had the largest total amount; an aspect of the results that stand out to me is that 2011 is significantly lower than the other fiscal years**

```{r}

pg_schools |>
  group_by(fiscal_year) |> 
  summarise(total_sum = sum(amount)) |>
  arrange(desc(total_sum))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

**A3: To clean the list to ensure that we only have one row per vendor and zip code, we would want to use the distinct() function (ie. for** PLANO-COUDON LLC, WHITING TURNER CO, & DISYS SOLUTIONS INC **they are all duplicated). To clean data that have nearly identical vendors, I'd rename them to be the same by shortening them using the "str_sub" function.**

```{r}

umd_capital <- cleaned_payments |> 
  filter(agency_name == "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND") |> 
  group_by(vendor_name, vendor_zip) |> 
  summarise(total_amount = sum(amount)) |> 
  arrange(desc(total_amount))
  
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

**A4: Between 2014-2018 and 2021, the total amount of money dipped below average and then in 2023, spiked up to a total amount that is almost equivalent to the starting year 2009.**

```{r}

cleaned_md_grants |> 
  filter(str_detect(description, "Stem Cell") | str_detect(description, "STEM CELL")) |> 
  group_by(fiscal_year) |> 
  summarise(total_amount = sum(amount))
  
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

**A5: There are 18 grantees shown in the list, but technically about 13 actual grantees due to the duplicates that are still showing up because they were inputed under different names. To clean it so that it was actually one record per grantee, I would shorten the different names so that they were the same (ie. "University of Maryland" becomes "UMD"). Grantees that are the most interesting to me are Haystack Solutions because it has the most money besides the University and UMD OFF OF THE COMPTROLLER because I'm not familiar with what this department is.**

```{r}

cleaned_md_grants |> 
  mutate(zip5 = str_sub(zip_code, start=1L, end=5L)) |>
  filter(zip5 == 20742) |>
  distinct() |>
  group_by(grantee) |> 
  select(grantor, grantee, zip5, amount) |> 
  summarise(total_amount = sum(amount))
```
