---
title: "week7_recap"
author: "Daniel Trielli"
date: "2023-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load Data

Run the codeblock below to load the data.

```{r}
earthquakes <- read_csv('https://earthquake.usgs.gov/earthquakes/feed/v1.0/summary/all_month.csv')

#Setting time column as datetime
earthquakes <- earthquakes |> mutate(time = as_datetime(time))
```

#### Answer the questions below

Most questions have a code block and a space for an answer below. Write the code you think is necessary and, in the answer space, write out what you did and what was the result.

------------------------------------------------------------------------

#### **Q1** Do you notice anything different with this read_csv()? Particularly in the file name?

**A1: It's a link to an active csv file instead of a csv file that is downloaded and added into the lab folder**

------------------------------------------------------------------------

#### **Q2** How many records there are there in this dataset? What do they mean and what useful information we can gather from it, looking at the columns?

**A2: 10,488 records but the records can change as it is a living dataset. The observations made in the dataset are earthquakes (or a detected tremor by the USGS). Some useful information that we can gather from looking at the columns are locations, severity, a close city, time and place.**

------------------------------------------------------------------------

#### **Q3** How do I reorganize this list to see the ones that are the deepest first? What is the depth that shows up for the deepest one, and its magnitude?

```{r}

earthquakes |> 
  arrange(desc(depth))

```

**A3: The depth that shows up as the deepest is Vanuatu region with 665.057 with a magnitude of 4.20 mb.**

------------------------------------------------------------------------

#### **Q4** I just want to see the earthquakes with a magnitude larger than 6. How do I do that? And how many are there that fit this criteria?

```{r}

earthquakes |> 
  filter(mag > 6)
```

**A4: There are 4 termor types that fit this criteria**

------------------------------------------------------------------------

#### **Q5** What about if I want to see earthquakes that have both a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [both]{.underline} these criteria?

```{r}

earthquakes |> 
  filter(mag > 6 & depth < 20)

```

**A5: Only one observation fits both criteria**

------------------------------------------------------------------------

#### **Q6** What about if I want to see earthquakes that either have a magnitude larger than 6 and a depth smaller than 20? How many are there in the data set that fit [either]{.underline} these criteria?

```{r}

earthquakes |> 
  filter(mag > 6 | depth < 20)
```

**A6: There are 8,247 observations that fit either criteria**

------------------------------------------------------------------------

#### **Q7** I'm interested in finding earthquakes that took place in Alaska. Which column should I look at? How do I use it to find all the earthquakes in Alaska? How many earthquakes took place there?

```{r}

earthquakes |> 
  filter(str_detect(locationSource, "ak"))

```

**A7: We should look at the "locationSource" column and str_detect so that you can filter for "ak" to find earthquakes that take place in Alaska. There are 2,336 earthquakes that took place in Alaska.**

------------------------------------------------------------------------

#### **Q8:** I notice that there is a column called 'type', that seems to have different kinds of tremors. What is the best way to find what are all the possible types of tremors, and counting how many of each there are in this data set? What are the first two most common types of tremors in this data set?

```{r}

earthquakes |> 
  group_by(type) |> 
  summarise(total_type = n()) |>
  arrange(desc(total_type))
```

**A8: The two most common types are earthquakes and quarry blasts**

------------------------------------------------------------------------

#### **Q9** How do I calculate the average depth of each type of tremor? What is the average depth of the earthquake type in this data set? Is there anything that seems unusual or surprising in this finding?

```{r}

earthquakes |> 
  group_by(type) |> 
  summarise(total_type = n(), avg_depth = mean(depth))
```

**A9: The average depth for an earthquake is 21.92. A thing that seems unusual about this outcome is that there are two negative average depths (found in explosion and quarry blast).**

------------------------------------------------------------------------

#### **Q10** I'm interested, in the future, to see the hours in which earthquakes happen. How can I extract the hour from the time column?

```{r}

earthquakes |> 
  mutate(hour = hour(time))

```

**A10: You can extract the hour by using the hour() function in mutate() with the time column.**

------------------------------------------------------------------------

#### **Q11** I want to make a note of all the records in this data set that I consider serious. For my purposes, I'm thinking that all tremors that have a magnitude that is larger than 3 are serious. How do I automatically create a new column making that distinction?

```{r}

earthquakes |> 
  mutate(mag_scale = case_when(mag >= 3 ~ "serious", mag <= 2 ~ "unserious")) |>
  select(mag_scale, mag) 
```

**A11: You would create a new column using mutate() and then use case_when() to create an expression that tells are to find magnitudes over 3 and label them as "serious" and then under 2 as "unserious" in the new mutated column.**

------------------------------------------------------------------------

#### **Q12** I have no idea how earthquakes work and I'm interested in seeing if there is a particular time of day in which serious earthquakes happen. How can I see that condensed in a table with all the hours in a day and all the serious earthquakes in each hour? What is the hour with fewer serious earthquakes and the one with the most serious earthquakes?

```{r}

earthquakes |> 
  filter(type == "earthquake") |> 
  mutate(hour = hour(time)) |> 
  group_by(hour, mag) |> 
  select(hour, mag) |> 
  arrange(desc(mag))

earthquakes |> 
  filter(type == "earthquake") |> 
  mutate(hour = hour(time)) |> 
  group_by(hour, mag) |> 
  select(hour, mag) |> 
  arrange(mag)
```

**A12**: The hours with the most serious earthquakes are 21, 14, 9, 11 & 16. The hours with the least serious earthquakes are 0, 20, 15, & 16.

------------------------------------------------------------------------

#### EXTRA Q What other questions are you interested in and how would you ask them in R?

1.  What's the largest earthquake in New Mexico?

```{r}

earthquakes |> 
  filter(locationSource == "nm") |>
  select(locationSource, mag, time) |> 
  arrange(desc(mag))
```

2.  How many earthquakes are happening by state?

```{r}

earthquakes |> 
  select(locationSource, type) |>
  group_by(locationSource) |> 
  summarise(total_earthquakes = n()) |> 
  arrange(desc(total_earthquakes))
```

3.  What is the average horizontal error for an earthquake?

```{r}

earthquakes |> 
  group_by(type) |> 
  drop_na(horizontalError) |> 
  summarise(total_type = n(), avg_horizontal_error= mean(horizontalError))
```
