---
title: "2024_EAVS_UOCAVA"
output: html_document
date: "2026-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(janitor)
```

## UOCAVA 2024 Data 

NOTES:

I created a dataset based on the Excel spreadsheet provided and then created a new column that calculated the rejection rate by county. A "NaN" value is showing up in the percentage column — idk what that is.

```{r}

uocava_2024 <- read_csv("eavs_2024_public_records.csv") 

uni_2024_voting <- uocava_2024 |> 
  clean_names() |>
  rename(ballots_cast = b11b, 
         ballots_rejected = b24b) |> 
  select(jurisdiction_name, state_abbr, ballots_cast, ballots_rejected) |> 
  mutate(across(everything(), ~ if_else(str_detect(.x, "Data not available"), NA, .x))) |> 
  mutate(across(everything(), ~ if_else(str_detect(.x, "Valid skip"), NA, .x))) |> 
  mutate(ballots_cast = as.numeric(ballots_cast), ballots_rejected = as.numeric(ballots_rejected)) |> 
  drop_na(ballots_cast, ballots_rejected) |> 
  mutate(pct_2024_rejected = (ballots_rejected/ballots_cast)*100)
```

NOTE: Arranged to see the top rejections by county (Bell County, TX has a high rejection rate of 13.13%)

```{r}

uni_2024_voting |> 
  arrange(desc(ballots_rejected))
```

## UOCAVA 2022 Data 

NOTES:

Basically did the same thing as above expect with 2022 EAVS data.

```{r}

uocava_2022 <- read_csv("eavs_2022_records.csv") 

uni_2022_voting <- uocava_2022 |> 
  clean_names() |>
  rename(ballots_cast = b9b, 
         ballots_rejected = b18b) |> 
  select(jurisdiction_name, state_abbr, ballots_cast, ballots_rejected) |> 
  mutate(across(everything(), ~ if_else(.x == -99, NA, .x))) |> 
  mutate(across(everything(), ~ if_else(.x == -88, NA, .x))) |> 
  mutate(across(everything(), ~ if_else(.x == -77, NA, .x))) |> 
  mutate(ballots_cast = as.numeric(ballots_cast), ballots_rejected = as.numeric(ballots_rejected)) |> 
  drop_na(ballots_cast, ballots_rejected) |> 
  mutate(pct_2022_rejected = (ballots_rejected/ballots_cast)*100)
```

```{r}

uni_2022_voting |> 
  arrange(desc(ballots_rejected))
```

## UOCAVA 2020 Data 

```{r}

uocava_2020 <- read_csv("eavs_2020_records.csv") 

uni_2020_voting <- uocava_2020 |> 
  clean_names() |>
  rename(ballots_cast = b9b, 
         ballots_rejected = b18b) |> 
  select(jurisdiction_name, state_abbr, ballots_cast, ballots_rejected) |> 
  mutate(across(everything(), ~ if_else(.x == -99, NA, .x))) |> 
  mutate(across(everything(), ~ if_else(.x == -88, NA, .x))) |> 
  mutate(across(everything(), ~ if_else(.x == -77, NA, .x))) |> 
  mutate(ballots_cast = as.numeric(ballots_cast), ballots_rejected = as.numeric(ballots_rejected)) |> 
  drop_na(ballots_cast, ballots_rejected) |> 
  mutate(pct_2020_rejected = (ballots_rejected/ballots_cast)*100)
```

```{r}


uni_2020_voting |> 
  arrange(desc(ballots_rejected)) 
```
